BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-01 implementation.tex
Callbacks Tracing Done

\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Log relative to a \texttt{NtOpenFile} call},captionpos=b]
~~2868~~ 1072 ntdll.dll!NtOpenFile
1072 	arg 0: 0x000fe810 (type=HANDLE*, size=0x4)
1072 	arg 1: 0x100020 (type=unsigned int, size=0x4)
1072 	arg 2: len=0x18, root=0x0, name=210/538 "\??\C:\Windows\WinSxS\
		x86_microsoft.windows.gdiplus_6595b64144ccf1df_1.1.7601.23894_none_5c0be957a009922e", 
		att=0x40, sd=0x00000000, sqos=0x00000000 (type=OBJECT_ATTRIBUTES*, size=0x4)
1072 	arg 3: 0x000fe7c0 (type=IO_STATUS_BLOCK*, size=0x4)
1072 	arg 4: FILE_SHARE_READ|FILE_SHARE_WRITE (type=named constant, value=0x3, size=0x4)
1072 	arg 5: FILE_DIRECTORY_FILE|FILE_SYNCHRONOUS_IO_NONALERT
 		(type=named constant, value=0x21, size=0x4)
1072    succeeded =>
1072 	arg 0: 0x000fe810 => 0x58 (type=HANDLE*, size=0x4)
1072 	arg 3: status=0x0, info=0x1 (type=IO_STATUS_BLOCK*, size=0x4)
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

The first thing which is done is determining whether or not the API under analysis has been invoked directly from the main executable of the Pin application. The concepts involved here are exactly the same as the ones applying to the main image check done for Native APIs, which was detailed in \textit{Section 3.3.1}. In short, it is checked if the API's return address, which is passed to the analysis function through the use of \texttt{IARG\_RETURN\_IP}, falls into the main executable memory address range, which was previously determined via \texttt{IMG\_HighAddress} and \texttt{IMG\_LowAddress}.

@@

The \texttt{apicall\_t struct} is initialized utilizing the API-related information, which was fetched beforehand during instrumentation and was passed to the analysis function via \texttt{libPointer}, as well as the values of the API arguments, which were provided to the analysis function through \texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE} as already discussed in \textit{Section 3.4.1}.

@@

This API allows to register a notification function to execute just before the pinned application changes context because of the reception of an asynchronous event, like for instance a callback or a Windows APC. In BlueTracer, the idea is to use such registered notification function to extract and log information relative to the context change. The notification function in question is of type \texttt{CONTEXT\_CHANGE\_CALLBACK} and the most important arguments it takes as input are:
\begin{itemize}
\item A member of the \texttt{CONTEXT\_CHANGE\_REASON} enumeration indicating the reason for the context change. This is the parameter allowing BlueTracer to distinguish between callbacks and APCs. In fact, in the case of callbacks it is set to \texttt{CONTEXT\_CHANGE\_REASON\_CALLBACK}, while in the case of APCs it is set to \texttt{CONTEXT\_CHANGE\_REASON\_APC}. 
\item The application's register state before the context change. It is \texttt{null} if the context change is caused by a callback.
\item The application's register state after the context change.  
\end{itemize}

Having outlined the tools made available by Pin to deal with the application's context change, let us now explain in detail how callbacks tracing and Windows APCs tracing is carried out in BlueTracer.

@@

To further clarify how the tracing of callbacks takes place in BlueTracer, let us show how the stack looks like right before the execution of \texttt{KiUserCallbackDispatcher}, i.e. when the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function is called (\textit{Figure 3.6}). Note that the return address is \texttt{null}.

It is worth mentioning that, in Win32, kernel mode to user mode callbacks are used exclusively by User32 for windowing related aspects. For this reason, in order for the kernelcallback table to not be null it has to be ensured at user32 is loaded. Moreover, it is also checked if callbacks belong to user32.dll.

The final output of the log looks like this:

>>>>

\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Log entry relative to a \texttt{NtOpenFile} call},captionpos=b]
~~2868~~ 1072 ntdll.dll!NtOpenFile
1072 	arg 0: 0x000fe810 (type=HANDLE*, size=0x4)
1072 	arg 1: 0x100020 (type=unsigned int, size=0x4)
1072 	arg 2: len=0x18, root=0x0, name=210/538 "\??\C:\Windows\WinSxS\
		x86_microsoft.windows.gdiplus_6595b64144ccf1df_1.1.7601.23894_none_5c0be957a009922e", 
		att=0x40, sd=0x00000000, sqos=0x00000000 (type=OBJECT_ATTRIBUTES*, size=0x4)
1072 	arg 3: 0x000fe7c0 (type=IO_STATUS_BLOCK*, size=0x4)
1072 	arg 4: FILE_SHARE_READ|FILE_SHARE_WRITE (type=named constant, value=0x3, size=0x4)
1072 	arg 5: FILE_DIRECTORY_FILE|FILE_SYNCHRONOUS_IO_NONALERT
 		(type=named constant, value=0x21, size=0x4)
1072    succeeded =>
1072 	arg 0: 0x000fe810 => 0x58 (type=HANDLE*, size=0x4)
1072 	arg 3: status=0x0, info=0x1 (type=IO_STATUS_BLOCK*, size=0x4)
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

The first thing which is done is determining whether or not the API under analysis has been invoked directly from the main executable of the Pin application. The concepts involved here are essentially the ones applying to the main image check done for Native APIs, which was detailed in \textit{Section 3.3.1}. In short, it is checked if the API's return address, which is passed to the analysis function through the use of \texttt{IARG\_RETURN\_IP}, falls into the main executable memory address range, which was previously determined via \texttt{IMG\_HighAddress} and \texttt{IMG\_LowAddress}.

@@

The \texttt{apicall\_t struct} is initialized utilizing the API-related data, which was fetched beforehand during instrumentation and was passed to the analysis function via \texttt{libPointer}, as well as the values of the API arguments, which were provided to the analysis function through \texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE} as already discussed in \textit{Section 3.4.1}.

@@

This API allows to register a notification function to execute just before the pinned application changes context due to the reception of an asynchronous event, like for instance a callback or a Windows APC. In BlueTracer, the idea is to use such registered notification function to extract and log information relative to the context change. The notification function in question is of type \texttt{CONTEXT\_CHANGE\_CALLBACK} and the most important arguments it takes as input are:
\begin{itemize}
\item A member of the \texttt{CONTEXT\_CHANGE\_REASON} enumeration indicating the reason for the context change. This is the parameter allowing BlueTracer to distinguish between callbacks and APCs. In fact, in the case of callbacks it is set to \texttt{CONTEXT\_CHANGE\_REASON\_CALLBACK}, while in the case of APCs it is set to \texttt{CONTEXT\_CHANGE\_REASON\_APC}. 
\item The application's register state before the context change. It is \texttt{null} if the context change is caused by a callback.
\item The application's register state after the context change.  
\end{itemize}

Having outlined the tools made available by Pin to deal with the application's context changes, let us now explain in detail how callbacks tracing and Windows APCs tracing is carried out in BlueTracer.

@@

Right before the execution of \texttt{KiUserCallbackDispatcher}, i.e. when the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function is called, the stack looks as follows (\textit{Figure 3.6}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/StackCallback.pdf}
\caption{\textit{Stack before \texttt{KiUserCallbackDispatcher} execution}}
\end{figure}

Having this in mind, it is now possible to explain step-by-step how the tracing of callbacks takes place in BlueTracer:

\begin{enumerate}
\item The stack pointer value is retrieved. This is achieved through the use of the \texttt{PIN\_GetContextReg} API, that allows to get the value of a register in the context provided as input. Such context, in our specific case, is the application's register state after the context change, which, as previously mentioned, is passed as argument to the \texttt{CONTEXT\_CHANGE\_CALLBACK} notification function.
\item The callback number is obtained by adding 4 to the stack pointer and fetching the pointed value, in accordance to \textit{Figure 3.6}.
\item The \texttt{KernelCallbackTable} is retrieved from the PEB. 
\item The function pointer stored in the \texttt{KernelCallbackTable} entry indicated by the callback number is acquired.
\item It is checked if the address of the function falls within \texttt{User32.dll} memory address range. In fact, in \texttt{Win32} user mode callbacks are employed exclusively by \texttt{User32.dll} for windowing related aspects \cite{UCDispatcher}, meaning that the retrieved function address must belong to \texttt{User32.dll}'s memory address range. Such range is determined by retrieving an handle for \texttt{User32.dll} through \texttt{GetModuleHandle}, building an \texttt{IMG} object from it utilizing Pin's \texttt{IMG\_FindByAddress} and invoking the usual \texttt{IMG\_LowAddress} and \texttt{IMG\_HighAddress} functions. 
\item The name of the callback is determined by creating a \texttt{RTN} object from the function address via \texttt{RTN\_FindByAddress} and consequently invoking \texttt{RTN\_Name} on such object.
\item The callback occurrence and the callback name is recorded in the log.
\end{enumerate}

The result of the aforementioned callback tracing procedure is a log entry having the following format (\textit{Listing 3.16}).
\\
\lstset{
    language=C++,
    frame=single, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Log entry relative to a user mode callback},captionpos=b]
Context change caused by callback -> USER32.dll!LoadMenuW
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
